Staged Abstract Interpreters

### 2019-09-14 paper/10_intro.tex
polish

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an operational
approach to constructing abstract interpreters.  By deriving semantics artifacts
for abstract interpretation from their concrete counterparts (for example,
abstract machines), soundness can be easily established by examining the
transformation of semantic artifacts. This systematic approach to abstraction
can be tailored to different language features (such as state, first-class
control, exceptions, concurrency, etc.) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
Darais:2015:GTM:2814270.2814308}. It has also been applied to various small-step
abstract machines \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
Sergey:2013:MAI:2491956.2491979} and big-step definitional interpreters
\cite{Wei:2018:RAA:3243631.3236800, DBLP:journals/pacmpl/DaraisLNH17,
Keidel:2018:CSP:3243631.3236767}.

%Based on the idea of abstracting abstract machines, more
%pragmatically, several implementation strategies utilizing purely
%functional programming techniques to build abstract interpreters have
%emerged.

Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows \cite{Keidel:2018:CSP:3243631.3236767},
extensible effects \cite{Kiselyov:2015:FMM:2804302.2804319, Githubsemantic},
etc. These pure approaches provide certain benefits, including the fact that the
abstract interpretation artifacts can be built in a compositional and modular
fashion. Furthermore, referential transparency allows programmers to confidently
reason about the correctness of their implementations.  The soundness of an
analysis can also be proven with less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of interpreted languages. For example, the list monad can be used to model
nondeterminism of collecting interpreters. Although such pure approaches have
their merits, they are significantly slower when compared with imperative,
stateful implementations.  In order to reconcile these approaches, one need to
use a programming technique which allows for high-level abstractions during
implementation, but bypasses the incurred overhead at runtime.

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. The first Futamura projection states that specializing an
interpreter to an input program removes interpretative overhead and yields an
equivalent executable that runs faster. In recent years, the idea of Futamura
projections has attracted more attention, and has been successfully applied to
many real-world problems, including PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, query compilers
\cite{DBLP:conf/sigmod/TahboubER18}, etc. The challenge here: can we apply
Futamura projections to a general abstract interpreter to make it run faster?
The solution is not obvious, as we need to specialize a nonstandard semantics
that extensively uses nondeterminism, operates on abstract domains, and
computes fixed-points iteratively.

@@

The first Futamura projection shows that specializing an
interpreter w.r.t. a input program yields an equivalent executable. For
instance, assume that @eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we can
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The first proposed approach to realizing Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes between compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program
over a two-level syntax \cite{Nielson:1992:TFL:130665} by manually annotating
the binding-times of variables. The MSP system will then check whether these
annotations are consistent, and specialize the program using that information.
The staging annotations can be either syntactic (e.g., quote and quasi-quote in
MetaML/MetaOCaml) or type-based (e.g., the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala).

@@

\paragraph{Contributions} Briefly, the contributions of the paper are as follows:
\begin{itemize}[leftmargin=2em]
  \item Intellectually, we present a framework that naturally extends the first
    Futamura projection to definitional abstract interpreters, showing a
    well-grounded approach to optimizing static analyses via meta-programming.
  \item Practically, we show that staging an abstract interpreter is useful to
    improve performance and scalability of analyses by case studies and an
    empirical evaluation.
\end{itemize}

>>>>

Recent progress on methodologies such as Abstracting Abstract Machines (AAM)
\cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10} uncovers an operational
approach to constructing abstract interpreters.  By deriving semantics artifacts
for abstract interpretation from their concrete counterparts (for example,
abstract machines), soundness can be easily established by examining the
transformation of semantic artifacts. This systematic approach to abstraction
can be tailored to different language features (such as state, first-class
control, exceptions, etc.) and sensitivity analyses
\cite{DBLP:conf/icfp/Gilray0M16, DBLP:conf/popl/GilrayL0MH16,
Darais:2015:GTM:2814270.2814308}. It has also been applied to various small-step
abstract machines \cite{DBLP:journals/jfp/HornM12, DBLP:conf/icfp/HornM10,
Sergey:2013:MAI:2491956.2491979} and big-step definitional interpreters
\cite{Wei:2018:RAA:3243631.3236800, DBLP:journals/pacmpl/DaraisLNH17,
Keidel:2018:CSP:3243631.3236767}.
Based on the idea of AAM, more pragmatically, several implementation strategies
in high-level functional programming languages have emerged.  Such techniques
include monads and monad transformers \cite{DBLP:journals/pacmpl/DaraisLNH17,
Sergey:2013:MAI:2491956.2491979}, arrows \cite{Keidel:2018:CSP:3243631.3236767},
extensible effects \cite{Githubsemantic},
etc. These pure approaches provide certain benefits, including the fact that the
abstract interpretation artifacts can be built in a compositional and modular
fashion. Furthermore, referential transparency allows programmers to confidently
reason about the correctness. The soundness of an
analysis can also be proven with less effort, whether by mechanized proofs
\cite{Darais:2016:CGC:2951913.2951934} or paper-based proofs
\cite{Keidel:2018:CSP:3243631.3236767}.

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs interpretive
overhead, such as pattern matching on the AST nodes and recursive calls on
subexpressions. If this requires only a single traversal, such overhead can be
negligible. Usually, however, abstract interpreters iteratively analyze and
traverse an AST multiple times to reach fixed-points; the accumulated overhead
is therefore magnified.
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).
Second, it is common that abstract interpreters written in pure languages make
extensive use of functional encodings of side-effects to implement the semantics
of interpreted languages. For example, the list monad can be used to model
nondeterminism of collecting interpreters. Although such pure approaches have
their merits, they are significantly slower when compared with imperative,
stateful implementations. The goal of this paper is to develop an approach for
implementing abstract interpreters, which allows high-level abstractions during
implementation for clarity, but passes the incurred overhead at runtime for
efficiency.
%In order to reconcile these approaches, one need to
%use a programming technique which allows for high-level abstractions during
%implementation, but bypasses the incurred overhead at runtime.

Roughly at the same time abstract interpretation was proposed,
\citet{futamura1971partial} observed a close connection between interpreters and
compilers through a hierarchy of specializations, known since then as the
Futamura projections. 
The first Futamura projection states that specializing an interpreter to an
input program removes interpretative overhead and yields an equivalent
executable that runs faster. Moreover, a procedure that can specialize an
interpreter to any input program is equivalent to a compiler.
In recent years, the idea of Futamura projections has attracted more attention,
and has been successfully applied to many real-world scenarios, such as
building JIT compilers for dynamic languages
\cite{Bolz:2009:TMP:1565824.1565827, Marr:2015:TVP:2814270.2814275} and building query
compilers \cite{DBLP:conf/sigmod/TahboubER18}.  The question here is can we
apply Futamura projections to an abstract interpreter and obtain a compiler
for abstract interpretation?
%apply Futamura projections to a general abstract interpreter to make it run faster? 
The solution is not obvious, as we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively.

@@

The first Futamura projection shows that specializing an interpreter w.r.t. an
input program yields an equivalent program. For instance, assume that 
@eval : Expr --> Input --> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization we
obtain a specialized program $\texttt{eval}_{\texttt{e0}}$ : @Input --> Value@.
%which is the so-called \textit{equivalent executable}.
% By definition of the interpreter, they produce the same result when applied
% with the argument $\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $.
The earlier approach to realizing Futamura projections is partial
evaluation (PE) \cite{DBLP:books/daglib/0072559}.
Offline PE relies on a separate binding-time analysis (\textit{static} or
\textit{dynamic}), and online PE identifies the binding-times during the
specialization.
The underlying representation of binding-times can be viewed as a
two-level semantics \cite{NIELSON1989117, NIELSON198859,
Nielson:1992:TFL:130665} that distinguishes between compile-time computation
and run-time computation.  However, it is hard to precisely analyze
binding-times given an arbitrary program. As an alternative approach,
multi-stage programming (MSP) \cite{taha1999multi, DBLP:conf/pepm/TahaS97}
allows programmers to annotate the binding-times of variables and control which
part of the program should be specialized.
At the syntactic level, traditional MSP languages (e.g., MetaML/MetaOCaml) use
quasi-quotations and escapes as annotations. However, the syntactic
quasi-quotation is a weak two-level semantics, and does not always preserve the
evaluation order of staged expressions.
%The MSP system will then check whether these
%annotations are consistent, and specialize the program using that information.
Besides the syntactic approach, staging annotations can also be purely
type-based.  One of the examples is the Lightweight Modular Staging framework
\cite{DBLP:conf/gpce/RompfO10} in Scala, which provides a stronger guarantee of
evaluation order by generating code in administrative normal form.

@@

\paragraph{Contributions} 
1) Intellectually, we present a framework that naturally extends the first
Futamura projection to abstract interpreters, showing a well-grounded approach
to optimizing static analyses via meta-programming.  2) Practically, we show
that staging an abstract interpreter is useful to improve performance and
scalability of analyses by case studies and an empirical evaluation.

### 2019-09-14 paper/20_prelim.tex
polish

We consider a call-by-value $\lambda$-calculus in direct-style, extended with
numbers, arithmetic, recursion, and conditionals. Other effects such as
assignments can also be supported readily. Since we are interested in
analyzing the dynamic behavior of programs, we elide the static semantics. We also
assume that input programs are well-typed and all variables are distinct. The
abstract syntax is as follows:
\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursion
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad as an argument and produce another monad. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

\paragraph{The Interpreter} We can now construct the semantics-agnostic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the AST
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, method @close@ that
lifts $\lambda$-terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because further evaluation may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}

>>>>

We consider a call-by-value $\lambda$-calculus in direct-style, extended with
numbers, arithmetic, recursion, and conditionals. Other effects such as
assignments can also be supported readily. Since we are interested in
analyzing the dynamic behavior of programs, we elide the precise static semantics. We
assume that programs are well-typed and variables are distinct. The
abstract syntax is as follows:
\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursion
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

@@

Note that here the monadic object @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad type as argument and produce another monad type. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

\paragraph{The Interpreter} We can now construct the generic interpreter
in monadic form, as shown in Figure \ref{fig:shared_int}. The basic idea
of generic interpretation is to traverse the AST
while maintaining the effects (such as environment and state updates).
Note that the interpreter is written in open-recursive style: it cannot
refer to itself directly. Instead, @eval@ takes an additional parameter @ev@ of
type @Expr => Ans@ that refers to itself. Consequently, method @close@ that
lifts $\lambda$-terms to closures and method @ap_clo@ that applies functions also
takes an extra @ev@, because further evaluation may happen inside them.
To close the open recursion, we use a fixed-point operator @fix@.
For concrete-interpretation instantiation, @fix@ works like the Y combinator;
for abstract-interpretation instantiation, it will instrument the interpreter
by memoizing @ev@'s inputs and outputs, which ensures the termination of
abstract interpretation.
Finally, there is a top-level wrapper @run@. The return type @Result@
depends on the kind of monads being used and is therefore also left abstract.
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans
  type Result; def run(e: Expr): Result
\end{lstlisting}
